---
title: "Getting started with flexstanr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with flexstanr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(flexstanr)
```

flexstanr gives a Stan-based R package **one interface** for fitting its models
through either [rstan](https://mc-stan.org/rstan/) or (optionally)
[cmdstanr](https://mc-stan.org/cmdstanr/). Your package supplies its own compiled
models; flexstanr resolves them at run time, so the same fitting code works
whichever backend is installed.

This vignette walks through wiring flexstanr into a host package and using it.

## Wiring it into your package

From the root of your Stan package, run the setup helper once:

```{r, eval = FALSE}
flexstanr::useFlexStanR()
```

This adds `flexstanr` (and `rstan`, the default backend) to your `Imports` and,
while flexstanr is still pre-CRAN, an interim `Remotes: ACCIDDA/flexstanr` entry
so `remotes` / `pak` can install it from GitHub. Once flexstanr is on CRAN, pass
`on_cran = TRUE` to skip the `Remotes` entry.

## Building sampler options

`stan_options()` collects and validates sampler arguments for the chosen
backend, forwarding them **verbatim** so a call feels native to that backend:

```{r}
opts <- stan_options(chains = 2, iter = 500, seed = 1)
str(opts)
```

Each backend has its own argument vocabulary, and mixing them is caught early
with a "did you mean" hint rather than failing deep inside the sampler:

```{r}
# `parallel_chains` is a cmdstanr word; the rstan backend rejects it.
try(stan_options(backend = "rstan", parallel_chains = 4))
```

## Fitting a model

`fit_model()` dispatches to the backend recorded on the options and resolves the
compiled model by name from your package. A host fitting one of its own models
needs no extra arguments; the calling package is detected automatically.

```{r, eval = FALSE}
# `"coverage"` is resolved from your package's stanmodels (rstan) or
# inst/stan/coverage.stan (cmdstanr).
fit <- fit_model(
  "coverage",
  dat_stan  = data_list,
  init      = init_list,
  stan_opts = opts
)
```

## Reading a fit

The `backend_*` accessors read a fitted object without your code needing to know
which backend produced it:

```{r, eval = FALSE}
# posterior draws as an iterations x chains x parameters array
draws <- backend_draws_array(fit)

# named parameters, matching rstan::extract()'s shape
post <- backend_extract(fit, pars = c("beta", "sigma"))

# guard against the degenerate "no draws" case before using a fit
stopifnot(backend_has_draws(fit))
```

Unrecognized objects pass through `backend_has_draws()` as if they carry draws,
so test doubles are left untouched:

```{r}
backend_has_draws(list())
```

## Choosing cmdstanr

Pass `backend = "cmdstanr"` to `stan_options()`. cmdstanr is optional and not on
CRAN, so install it separately (see the cmdstanr
[getting-started guide](https://mc-stan.org/cmdstanr/)); selecting it without the
package installed errors early with an actionable message.

```{r, eval = FALSE}
opts <- stan_options(backend = "cmdstanr", parallel_chains = 4, iter_warmup = 500)
```
