---
title: "Parallel and threaded fitting (including HPC)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel and threaded fitting (including HPC)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
library(flexstanr)
```

Fitting is the expensive part of most Stan analyses. This article explains how
flexstanr splits your available cores between running chains in parallel and
threading within each chain, and how to wire that split into your package's fit
function so it runs efficiently on a laptop and on a scheduled HPC cluster.

The pieces are two functions -- `optimal_alloc()`, which decides the split, and
`configure_threading()`, which applies it to a `stan_options()` object for
whichever backend you use -- plus a one-line caveat about compiling your model
for threading.

## Two levels of parallelism

A fit can use cores two ways at once:

1. **Across chains.** MCMC chains are independent, so they run in parallel with
   near-linear speedup and no coordination overhead.
2. **Within a chain (threading).** If your Stan model sums a
   conditionally-independent term with `reduce_sum()` (or `map_rect()`), each
   chain can split that sum across threads.

Because chain-parallelism is "free" (no threading overhead) while within-chain
threading scales sub-linearly, `optimal_alloc()` **fills chain-parallelism first**
and gives leftover cores to threads:

| chains | cores | chains in parallel | threads per chain |
|-------:|------:|-------------------:|------------------:|
| 4      | 4     | 4                  | 1 (no spare)      |
| 4      | 8     | 4                  | 2                 |
| 4      | 16    | 4                  | 4                 |
| 2      | 8     | 2                  | 4                 |

So threading only "switches on" when you have more cores than chains -- otherwise
every core is spent running chains.

```{r}
optimal_alloc(4, cores = 8)
#> $parallel_chains
#> [1] 4
#> $threads_per_chain
#> [1] 2
```

## Wiring it into your fit function

`optimal_alloc()` returns the split; `configure_threading()` writes it onto your
`stan_options()` using the backend's native controls. The two backends receive
threads differently, and `configure_threading()` hides that:

* **rstan** runs chains in parallel via `cores` and reads the per-chain thread
  count from the `STAN_NUM_THREADS` environment variable at run time (it has no
  threads-per-chain argument). So `configure_threading()` sets `cores` and, as a
  side effect, `STAN_NUM_THREADS`.
* **cmdstanr** takes both natively, so it sets `parallel_chains` and
  `threads_per_chain` and touches no environment variable.

Because the rstan branch mutates the process environment, save and restore
`STAN_NUM_THREADS` around the fit so a run does not leak its thread count into
the rest of the session:

```{r}
run_my_fit <- function(dat_stan, init, stan_opts = stan_options(chains = 4)) {
  alloc <- optimal_alloc(stan_opts$chains)

  # rstan sets STAN_NUM_THREADS as a global side effect; restore it afterwards.
  old <- Sys.getenv("STAN_NUM_THREADS", unset = NA)
  on.exit(
    if (is.na(old)) Sys.unsetenv("STAN_NUM_THREADS")
    else Sys.setenv(STAN_NUM_THREADS = old),
    add = TRUE
  )

  stan_opts <- configure_threading(stan_opts, alloc)
  fit_model("my_model", dat_stan = dat_stan, init = init, stan_opts = stan_opts)
}
```

`fit_model()` forwards `cores` / `parallel_chains` / `threads_per_chain`
straight to the backend's sampler, so nothing else is needed at the call site.

### Compile your model for threading

Within-chain threads only do anything if the model was **compiled with threading
enabled**:

* **cmdstanr** -- `fit_model()` compiles with `cpp_options = list(stan_threads =
  TRUE)` automatically whenever the options carry `threads_per_chain > 1`, so
  there is nothing to do.
* **rstan** -- the host package's models must have been built with threading
  support (this is a package-build concern; models compiled by `rstantools`
  with the Stan threading backend read `STAN_NUM_THREADS`). If a model was not
  built for threading, setting `STAN_NUM_THREADS` simply has no effect.

## How flexstanr decides how many cores to use

When you don't pass `cores`, `optimal_alloc()` asks
[`parallelly::availableCores()`](https://parallelly.futureverse.org/) for the
number of cores the **process is allowed** to use -- crucially **not**
`parallel::detectCores()`. It respects

* the HPC scheduler's allocation (`SLURM_CPUS_PER_TASK`, PBS, SGE, LSF),
* cgroup CPU quotas (containers, cpuset),
* `options(mc.cores = ...)`,

and returns 2 under `R CMD check`. This makes the automatic split safe on a
shared cluster. You can always take control -- an explicit core count is used
exactly as given.

## On a laptop or workstation

Nothing to do: the default detects your cores and splits them. To cap usage, set
`mc.cores`:

```{r}
options(mc.cores = 6)
# with chains = 4: 4 chains in parallel, 1 thread each (6 %/% 4 = 1);
# 2 cores are left free.
```

## On an HPC cluster

### The trap to avoid

`parallel::detectCores()` reports **every** core on the compute node (often 64,
128+) regardless of your job's allocation. A fit sizing itself from that would
launch dozens of threads inside a job pinned to a handful of cores; the
scheduler's cgroup would throttle it to a crawl (or you would trample other
users on the node). Using `parallelly::availableCores()` reads your *allocation*
instead, which is why `optimal_alloc()` uses it.

### Single node: let flexstanr split your allocation

Request one task with several CPUs and let the automatic split divide them. For
4 chains x 2 threads (8 cores):

```bash
#!/bin/bash
#SBATCH --job-name=stan-fit
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=8      # availableCores() will report 8
#SBATCH --mem=16G
#SBATCH --time=04:00:00
Rscript fit.R
```

```{r}
# fit.R
fit <- run_my_fit(dat_stan, init, stan_opts = stan_options(chains = 4, iter = 2000))
saveRDS(fit, "fit.rds")
```

No core arithmetic in your script: `availableCores()` picks up
`SLURM_CPUS_PER_TASK = 8`, and the fit runs 4 chains x 2 threads.

### Many chains across nodes: one chain per array task

Chain-parallelism and threads are **shared-memory**: they cannot span nodes. To
scale beyond a single node, run **one chain per job** with a SLURM array, give
each job all its cores for threading, and combine the single-chain fits
afterwards.

```bash
#SBATCH --array=1-4            # one task per chain
#SBATCH --nodes=1
#SBATCH --cpus-per-task=8      # all 8 cores -> threads for this one chain
```

```{r}
# each array task fits a single chain with a distinct seed
i <- Sys.getenv("SLURM_ARRAY_TASK_ID")
fit <- run_my_fit(dat_stan, init, stan_opts = stan_options(chains = 1, seed = as.integer(i)))
saveRDS(fit, sprintf("chain-%s.rds", i))
```

## Choosing the thread count yourself

`optimal_alloc()` auto-splits, but you can override `threads_per_chain` on the
result before passing it to `configure_threading()`:

```{r}
alloc <- optimal_alloc(4)     # auto-detect cores
alloc$threads_per_chain <- 1L # one thread per chain (chains still parallel)

alloc <- optimal_alloc(4)
alloc$threads_per_chain <- 2L # exactly 2 threads per chain, whatever the machine
```

`optimal_alloc()` is always called for `parallel_chains` (how many chains run in
parallel, from the available cores); only `threads_per_chain` is overridden.
Setting more threads than a machine has free cores per chain will oversubscribe
-- that is your explicit choice, honored rather than clamped.

Parallelism is applied *after* `stan_options()`, not inside it:
`configure_threading()` writes each backend's parallelism fields (`cores` for
rstan; `parallel_chains` / `threads_per_chain` for cmdstanr) directly.
`stan_options()` is for how the sampler explores (`iter`, `warmup`, `seed`,
`control = list(adapt_delta = ...)`, ...).
